Loop-Aware Optimizations in PyPy's Tracing JIT

### 2012-08-16 talk/dls2012/paper.tex
date of luajit head
move stuff around

The loop peeling optimization was implemented in the PyPy
framework in about 450 lines of RPython code. That means that the JIT-compilers generated for all
interpreters implemented with RPython now can take advantage of
it. Benchmarks have been executed for a few different interpreters and
we see improvements in several cases. The ideal loop for this optimization
is short and contains numerical calculations with no failing guards and no
external calls. Larger loops involving many operations on complex objects
typically benefit less from it. Loop peeling never makes the generated code worse, in
the worst case the peeled loop is exactly the same as the preamble. 
Therefore we
chose to present benchmarks of small numeric kernels where loop peeling can show
its use.

@@

\subsection{Python}
The Python interpreter of the RPython framework is a complete Python
version 2.7 compatible interpreter. A set of numerical
calculations were implemented in both Python, C and Lua and their
runtimes are compared in Figure~\ref{fig:benchmarks}.\footnote{
    The benchmarks and the scripts to run them can be found in the repository for this paper:
    \texttt{https://bitbucket.org/pypy/extradoc/src/ tip/talk/dls2012/benchmarks}
}
For benchmarks using larger Python applications the times are unaffected or
slightly improved by the loop optimization of this paper.

@@

initiated to $x_0=y/2$ and the benchmark consists of a single loop updating this

@@

\begin{itemize}
\item {\bf SOR}$\left(n, c\right)$: Jacobi successive over-relaxation on a $n\times n$ grid repreated $c$ times.
The same custom two-dimensional array class as described above is used to represent
the gird.
\item {\bf SparseMatMult}$\left(n, z, c\right)$: Matrix multiplication between a $n\times n$ sparse matrix,
stored in compressed-row format, and a full storage vector, stored in a normal array. The matrix has $z$ non-zero elements and the calculation is repeated $c$ times.
\item {\bf MonteCarlo}$\left(n\right)$: Monte Carlo integration by generating $n$ points uniformly distributed over the unit square and computing the ratio of those within the unit circle.
\item {\bf LU}$\left(n, c\right)$: LU factorization of an $n \times n$ matrix. The rows of the matrix is shuffled which makes the previously used two-dimensional array class unsuitable. Instead a list of arrays is used to represent the matrix. The calculation is repeated $c$ times.
\item {\bf FFT}$\left(n, c\right)$: Fast Fourier Transform of a vector with $n$ elements, represented as an array, repeated $c$ times.
\end{itemize}

@@

\begin{itemize}
\item PyPy 1.9
\item CPython 2.7.1
\item GCC 4.5.2 shipped with Ubuntu 11.4
\item LuaJIT 2.0 beta, commit ID 0dd175d9e711f039c663d35e96c149b705bcf450
\end{itemize}

@@

Other interesting interpreters that are helped greatly by this optimization are
for example our Prolog interpreter written in
RPython~\cite{bolz_towards_2010}. Prolog programs often contain
tight
loops that perform list processing. Furthermore we experimented with a Python library
for writing numerical kernels doing array manipulation. The exact extent is
out of scope for this paper.

>>>>

The loop peeling optimization was implemented in RPython's tracing JIT
in about 450 lines of RPython code. That means that the JIT-compilers generated for all
interpreters implemented with RPython now can take advantage of
it. Benchmarks have been executed for a few different interpreters and
we see improvements in several cases.

An example of an RPython interpreter that is helped greatly by this
optimization is our Prolog interpreter~\cite{bolz_towards_2010}. Prolog
programs often contain tight loops that perform for example list processing.
Furthermore we experimented with a Python library for writing numerical kernels
doing array manipulation.

The ideal loop for this optimization
is short and contains numerical calculations with no failing guards and no
external calls. Larger loops involving many operations on complex objects
typically benefit less from it. Loop peeling never makes the generated code worse, in
the worst case the peeled loop is exactly the same as the preamble. 
Therefore we
chose to present benchmarks of small numeric kernels where loop peeling can show
its use.

@@

The Python interpreter of the RPython framework is a complete Python
version 2.7 compatible interpreter. A set of numerical
calculations were implemented in both Python, C and Lua and their
runtimes are compared in Figure~\ref{fig:benchmarks}.\footnote{
    The benchmarks and the scripts to run them can be found in the repository for this paper:
    \texttt{https://bitbucket.org/pypy/extradoc/src/ tip/talk/dls2012/benchmarks}
}
For benchmarks using larger Python applications the times are unaffected or
slightly improved by the loop optimization of this paper.

@@

initialized to $x_0=y/2$ and the benchmark consists of a single loop updating this

@@

\begin{itemize}
\item {\bf SOR}$\left(n, c\right)$: Jacobi successive over-relaxation on a $n\times n$ grid repreated $c$ times.
The same custom two-dimensional array class as described above is used to represent
the grid.
\item {\bf SparseMatMult}$\left(n, z, c\right)$: Matrix multiplication between a $n\times n$ sparse matrix,
stored in compressed-row format, and a full storage vector, stored in a normal array. The matrix has $z$ non-zero elements and the calculation is repeated $c$ times.
\item {\bf MonteCarlo}$\left(n\right)$: Monte Carlo integration by generating $n$ points uniformly distributed over the unit square and computing the ratio of those within the unit circle.
\item {\bf LU}$\left(n, c\right)$: LU factorization of an $n \times n$ matrix. The rows of the matrix is shuffled which makes the previously used two-dimensional array class unsuitable. Instead a list of arrays is used to represent the matrix. The calculation is repeated $c$ times.
\item {\bf FFT}$\left(n, c\right)$: Fast Fourier Transform of a vector with $n$ elements, represented as an array, repeated $c$ times.
\end{itemize}

@@

\begin{itemize}
\item PyPy 1.9
\item CPython 2.7.1
\item GCC 4.5.2 shipped with Ubuntu 11.4
\item LuaJIT 2.0 beta, git head of August 15, 2012, commit ID 0dd175d9e711f039c663d35e96c149b705bcf450
\end{itemize}

@@
